---
title: "Untitled"
author: "Christina Kim"
date: "9/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache.lazy = FALSE)

library("tidycensus")
library("tidyverse")
library('here')
library('janitor')

census_api_key("4d304bfde4c98d6bf6a672d2721b858f7307569b")
```

## Import Census Population Data

Data downloaded from the Census website at https://data.census.gov/cedsci/table?q=Housing%20Units%20in%20the%20United%20States&t=Housing%3AHousing%20Units%3APopulations%20and%20People%3ARace%20and%20Ethnicity&g=0400000US06.050000&y=2019&tid=ACSDP1Y2019.DP05&tp=true&hidePreview=true

```{r import_data}
data <- read.csv(here("data/census-demographics", "ACSDP1Y2019.DP05_data_with_overlays_2020-09-26T140825.csv"), header = TRUE)
```

Original data set has two rows of column headings, one for the codes, one for more readable names.

```{r review_data}
dim(data)
colnames(data)
str(data)
View(data)
```

Import the data set but skip the first row of code values.

```{r reimport_data_skip_1}
data_skip1 <- read_csv(here("data/census-demographics", "ACSDP1Y2019.DP05_data_with_overlays_2020-09-26T140825.csv"), skip = 1) %>% 
  clean_names()

data_skip1 %>% 
  colnames()
```

Clean up the data to rename columns and select only potentially useful columns.

Note: The 1977 OMB Race and Ethnic standards maintain that ethnicity (“Hispanic” or “Not Hispanic”) is a separate and distinct concept from race (“White,” “Black,” “American Indian or Alaskan Native,” or “Asian or Pacific Islander”.) Therefore individuals who are Hispanic may be of any Race. https://www2.census.gov/about/training-workshops/2018/2018-01-24-clmso-pres.pdf?

```{r}
data_skip1 %>% 
  # Remove margin of error values
  select(-contains("margin_of_error")) %>% 
  select(id, geographic_area_name,
         -contains("_1"),
         # Remove age fields
         !contains("_years") &
         !contains("voting_age") &
         # Include sex information
         matches("sex_and_age_.*male$") |
         # Include the Hispanic of Latino category for race and ethnicity (i.e., drop the "Race" category)
         ends_with("hispanic_or_latino_of_any_race") |
         ends_with("_alone") |
         ends_with("not_hispanic_or_latino_two_or_more_races"),
         # Remove sum rows (duplicate of other more detailed rows)
         # !contains("_1"),
         # Include housing unit count information in case it's helpful for housing analysis
         estimate_total_housing_units
         ) %>% 
  # Clean up the county column
  rename(county = geographic_area_name) %>% 
  mutate(county = str_remove(county, ", California")) %>% 
  filter(!county == "United States") %>% 
  rename_with(~ str_remove_all(.x, "(_sex_and_age|_hispanic_or_latino_and_race)_total_population|_alone|_of_any_race|_not_hispanic_or_latino|_or_african_american") %>% 
                str_replace("american_indian_and_alaska_native", "native_american") %>% 
                str_replace("native_hawaiian_and_other_pacific_islander", "pacific_islander")
              ) %>%
  mutate(estimate_other_races = estimate_some_other_race + estimate_two_or_more_races,
         percent_other_races = percent_some_other_race + percent_two_or_more_races,
         .before = estimate_total_housing_units) %>% 
  mutate(across(contains("percent"), ~ .x / 100)) %>% 
  select(-contains("some_other_race"), -contains("two_or_more_races")) %>% 
  write_csv(here("data/census-demographics", "county_sex_race_housing_units.csv"))


```

## Import Census Housing Data

Data downloaded from the Census website at https://data.census.gov/cedsci/table?q=housing%20in%20california&t=Housing%3AHousing%20Units&g=0400000US06.050000&y=2019&tid=ACSDP1Y2019.DP04&moe=false&tp=true&hidePreview=true

```{r import_data_housing}
data_housing <- read_csv(here("data/census-housing", "ACSDP1Y2019.DP04_data_with_overlays_2020-10-02T194057.csv"), skip = 1) %>% 
  clean_names()

data_housing
```

Original data set has two rows of column headings, one for the codes, one for more readable names.

```{r review_data_housing}
dim(data_housing)
colnames(data_housing %>% select(ends_with("housing_units")))
```


```{r clean_export_data_housing}
data_housing %>% 
  # Remove margin of error values
  select(-matches("margin_of_error")) %>% 
  # Remove duplicate sum columns
  select(-ends_with("_total_housing_units")) %>% 
  # Remove percent columns wihtout actual values - shown as "(X)"
  select(-(starts_with("percent") & where(is.character))) %>%
  # Choose columns
  select(id, geographic_area_name, matches("housing_occupancy_total_housing_units|vacancy_rate|units_in_structure|year_structure_built|median_rooms|occupants_per_room")) %>% 
  # Clean up the county column
  rename(county = geographic_area_name) %>% 
  mutate(county = str_remove(county, ", California")) %>% 
  filter(!county == "United States") %>% 
  mutate(across(contains("percent"), ~ .x / 100)) %>% 
  write_csv(here("data/census-housing", "census_housing_data_cleaned.csv"))
```

## CDC Wonder Race and Ethnicity Data

Data from: https://wonder.cdc.gov/bridged-race-v2019.html

Query Criteria:
Ethnicity:	Hispanic or Latino; Not Hispanic or Latino
Gender:	Female; Male
Race:	American Indian or Alaska Native; Asian or Pacific Islander; Black or African American; White
States:	California (06)
Yearly July 1st Estimates:	2019
Group By:	County; Gender; Ethnicity; Race
Show Totals:	False
Show Zero Values:	True
Data Table:	Default


```{r}
data_cdc <- read_tsv(here("data/cdc-demographics", "Bridged-Race Population Estimates 2019.txt")) %>% 
  clean_names()

data_cdc %>% View()
```

```{r}
data_cdc_trimmed <- data_cdc %>% 
  # Remove notes
  select(-notes) %>% 
  drop_na() %>% 
  # Clean up the county name
  mutate(county = str_remove(county, ", CA"))
```

```{r cdc_gender}
data_cdc_gender <- data_cdc_trimmed %>% 
  group_by(county, county_code, gender, gender_code) %>% 
  summarize(population = sum(population)) %>% 
  pivot_wider(id_cols = c(county, county_code), names_from = gender, values_from = population)
```

Categorize group as Hispanic or Latino as so in ethnicity/race, otherwise use the race column value.

```{r cdc_ethnicity_race}
data_cdc_race <- data_cdc_trimmed %>% 
  select(county, county_code, ethnicity, race, population) %>% 
  mutate(ethnicity_race = if_else(ethnicity == "Hispanic or Latino", "Hispanic or Latino", race)) %>% 
  group_by(county, county_code, ethnicity_race) %>% 
  summarize(population = sum(population)) %>% 
  select(county, county_code, ethnicity_race, population) %>% 
  pivot_wider(id_cols = c(county, county_code), names_from = ethnicity_race, values_from = population)
```


```{r}
data_cdc_all <- read_tsv(here("data/cdc-demographics", "All Bridged-Race Population Estimates 2019.txt")) %>% 
  clean_names()

data_cdc_all_trimmed <- data_cdc_all %>% 
  # Remove notes
  select(-notes) %>% 
  drop_na() %>% 
  # Clean up the county name
  mutate(county = str_remove(county, ", CA")) %>% 
  rename(total_population = population)
```

Merge all CDC clean data.

```{r}
data_cdc_all_trimmed %>% 
  left_join(data_cdc_gender, by = c("county", "county_code")) %>% 
  left_join(data_cdc_race, by = c("county", "county_code")) %>% 
  # Create an ID column that matches the census data
  mutate(id = paste0("0500000US", county_code), .before = county_code) %>% 
  clean_names() %>% 
  write_csv(here("data/cdc-demographics", "cdc_race_gender_population_data_cleaned.csv"))
```

## Binary Categorization by County

Using the Census ethnicity variable, 12 counties are dominantly Hispanic or Latino, while 46 counties are dominantly non-Hispanic or Latino.

```{r categorize_ethnicity}
data_cdc_ethnicity <- data_cdc_trimmed %>% 
  group_by(county, county_code) %>% 
  mutate(percent_population = population / sum(population)) %>% 
  group_by(county, county_code, ethnicity) %>% 
  summarise(percent_population = sum(percent_population)) %>% 
  pivot_wider(c(county, county_code), names_from = ethnicity, values_from = percent_population) %>% 
  clean_names() %>% 
  mutate(dominant_ethnicity = if_else(hispanic_or_latino >= not_hispanic_or_latino, "Hispanic or Latino", "Not Hispanic or Latino")) %>% 
  ungroup()
  
data_cdc_ethnicity %>% 
  count(dominant_ethnicity)

data_cdc_ethnicity
```

Using the Census race variable (i.e., not taking ethnicity into account), every county in California is dominantly white.

```{r categorize_race}
data_cdc_race_white <-
  data_cdc_trimmed %>% 
  group_by(county, county_code) %>% 
  mutate(percent_population = population / sum(population)) %>% 
  group_by(county, county_code, race) %>% 
  summarise(percent_population = sum(percent_population)) %>% 
  pivot_wider(c(county, county_code), names_from = race, values_from = percent_population) %>% 
  clean_names() %>% 
  rowwise() %>%
  mutate(dominant_race_white = if_else(max(c_across(american_indian_or_alaska_native:white)) == white, "Dominantly white", "Dominantly non-white")) %>% 
  ungroup()
  
data_cdc_race_white %>% 
  count(dominant_race_white)
```

Following the earlier logic of counting the ethnicity group first (i.e., Hispanic or Latino vs. not), and assigning non-Hispanic or Latino population into race groups after, 40 counties are dominantly white (non-Hispanic white) while 18 counties are dominantly non-white (Hispanics and non-Hispanics who do not identify as white).

Note this categorization counts only non-Hispanic white population as white population.

```{r categorize_race_ethnicity_original_groups}
data_cdc_race_hispanic_white <- data_cdc_race %>% 
  # Create an ID column that matches the census data
  mutate(id = paste0("0500000US", county_code), .before = county_code) %>% 
  clean_names() %>% 
  rowwise() %>%
  mutate(dominant_race_white = if_else(max(c_across(american_indian_or_alaska_native:white)) == white, "Dominantly white (non-Hispanic only)", "Dominantly non-white")) %>% 
  ungroup()
  
data_cdc_race_hispanic_white %>% 
  count(dominant_race_white)
```


```{r}
data_cdc_race_hispanic_white %>% 
  select(county, id, county_code, dominant_race = dominant_race_white) %>% 
  left_join(data_cdc_ethnicity %>% select(county, dominant_ethnicity)) %>% 
  write_csv(here("data/cdc-demographics", "cdc_race_gender_population_data_categorized.csv"))
```


## Tidy Census Analysis

Exploration only - decided not to use the package due to lack of 2019 data and not enough documentation for ACS data.

```{r}
census_vars <- load_variables(2010, "sf1", cache = TRUE)

View(census_vars)
```

The API call for the ACS data returns an error for year 2019.

```{r}
ca <- get_acs(geography = "county",
              variables = c(medincome = "B19013_001"),
              state = "CA",
              year = 2018)

ca
```

How does the example in the package readme categorize race and ethnicity?

Looking up the code P005003, it validates the example only picks out "Not Hispanic or Latino - White alone" combination as "white" and so on. https://api.census.gov/data/2010/dec/sf1/variables.html

```{r}
racevars <- c(White = "P005003",
              Black = "P005004",
              Asian = "P005006",
              Hispanic = "P004003")

counties <- get_decennial(geography = "county", variables = racevars,
                  state = "CA",
                  year = 2010,
                  summary_var = "P001001")

head(counties)
```

