---
title: 'Exploratory Data Analysis: Demographics, Housing and COVID-19'
author: "Christina Kim"
date: "10/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache.lazy = FALSE)

library("tidyverse")
library('here')
library("scales")
library("RColorBrewer") # Use display.brewer.all() to see all colors
library("wesanderson")
library("GGally")
library("janitor")
library("ggcorrplot")
```

## Summary Data Review

```{r import_data}
data <- read.csv(here("data", "summary_data.csv"))
head(data, 10)
```

```{r}
data_race_categorized <- read.csv(here("data/cdc-demographics", "cdc_race_gender_population_data_categorized.csv"))

df <- data %>% 
  left_join(data_race_categorized %>% select(-county, -id), by = c("countyid" = "county_code"), keep = FALSE) %>% 
  clean_names()

colnames(df)
```


There is a slight discrepancy between 2019 and 2020 population numbers but generally they are in the same ballpark.

```{r}
df %>% 
  select(countyname, county_pop2019, county_pop_2020) %>% 
  head(10)
```

County population varies quite a bit with Los Angeles county being much larger in population than any other counties in the state.

```{r}
df %>% 
  ggplot(aes(x = reorder(countyname, county_pop2019), y = county_pop2019)) +
  geom_bar(stat = "identity", fill = wes_palette(n = 1, "GrandBudapest1")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title="Large variance in population between counties", subtitle="Los Angeles county has more poulation than any other counties in CA", x="Counties in California", y = " Population") +
  coord_flip() +
  theme_minimal()
ggsave(here("output", "population.png"), width = 10, height = 10)
```


### What is the demographics makeup in California?

```{r}
df_demos <- df %>% 
  select(countyname, county_pop2019, percent_female:percent_white) 

# Double check percentage add up to 100%
df_demos %>% 
  rowwise() %>% 
  transmute(countyname, sum_races = sum(c_across(percent_american_indian_or_alaska_native:percent_white)))
```

With the exception of a handful of counties, the majority have a roughly even split between percent male and percent female residents.

```{r}
# Visualize sex percentages
df_demos %>% 
  select(countyname, matches("percent_(fe)?male")) %>% 
  pivot_longer(c(percent_female, percent_male), names_to = "sex") %>% 
  ggplot(aes(x = reorder(countyname, desc(countyname)), y = value, fill = sex)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = wes_palette(n = 2, name = "GrandBudapest1"),
                    name = "Sex",
                    labels = c("Female", "Male")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title="Majority of counties have an even sex split", subtitle="Percentage female vs. percentage male by county", x="Counties in California", y = "% Population") +
  coord_flip() +
  theme_minimal()
ggsave(here("output", "population_sex.png"), width = 10, height = 10)
```

On the other hand, there is a large variance in race and ethnicity makeups by county. With White and Hispanic/Latino groups being the dominant group in a county most often, a handful of counties have a notable dominance by the Native American or Asian group. Black group is a minority in all counties.

```{r}
# Visualize race percentages
df_demos %>% 
  select(countyname, contains("percent") & !matches("percent_(fe)?male")) %>% 
  pivot_longer(contains("percent"), names_to = "race_ethnicity") %>% 
  ggplot(aes(x = reorder(countyname, desc(countyname)), y = value, fill = race_ethnicity)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = wes_palette(n = 5, name = "Darjeeling1"),
                    name = "Race and Ethnicity",
                    labels = c("Native American", "Asian or Pacific Islander", "Black", "Hispanic or Latino", "White")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title="Large variance in race and ethnicity makeup by county", subtitle="Percentage of each race and ethnicity group by county", x="Counties in California", y = "% Population") +
  coord_flip() +
  theme_minimal()
ggsave(here("output", "population_race.png"), width = 10, height = 10)
```

```{r}
df %>% 
  drop_na(contains("income")) %>% 
  ggplot(aes(x = reorder(countyname, median_household_income_in_dollar), y = median_household_income_in_dollar)) + 
  geom_bar(stat = "identity", fill = wes_palette(n = 1, "GrandBudapest1")) +
  scale_y_continuous(labels = scales::dollar) +
  coord_flip() +
  labs(title="Household income varies a lot from county to county", subtitle="The highest median income is almost 2x the lowest", x="Counties in California", y = "Median Household Income") +
  theme_minimal()
ggsave(here("output", "income.png"), width = 10, height = 10)
```


### How does the housing situation vary by county?

Note: The housing data is not available for all counties, and the counties without data were omitted.

```{r}
df %>% 
  drop_na() %>% 
  ggplot(aes(x = reorder(countyname, total_housing_units), y = total_housing_units)) +
  geom_bar(stat = "identity", fill = wes_palette(n = 1, "GrandBudapest1")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title="Large variance in total housing units between counties", subtitle="Los Angeles county has more poulation than any other counties in CA", x="Counties in California", y = " Housing Units") +
  coord_flip() +
  theme_minimal()
ggsave(here("output", "housing_units.png"), width = 10, height = 10)
```

```{r}
df %>% 
  ggplot(aes(x = county_pop2019, y = total_housing_units)) +
  geom_point() +
  labs(title="Direct correlation between population and available housing units per county", subtitle="", x="Population", y = "Total Housing Units") +
  theme_minimal()
ggsave(here("output", "population_housing.png"), width = 10, height = 10)
```


```{r}
df %>% 
  drop_na() %>% 
  ggplot(aes(x = reorder(countyname, estimate_rooms_total_housing_units_median_rooms), y = estimate_rooms_total_housing_units_median_rooms)) +
  geom_bar(stat = "identity", fill = wes_palette(n = 1, "GrandBudapest1")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title="Median number of rooms is the lowest in metropolitan areas", subtitle="SF, LA and Alameda counties have a much lower number of rooms (more crowded)", x="Counties in California", y = " Median Number of Rooms per Housing") +
  coord_flip() +
  theme_minimal()
ggsave(here("output", "housing_median_rooms.png"), width = 10, height = 10)
```

Majority of the housing units are relatively new across most of the counties. San Francisco and LA counties, however, have relatively more older housing units than other counties.

```{r}
# Visualize housing ages
df %>% 
  select(countyname, percent_old, percent_new) %>% 
  drop_na() %>% 
  pivot_longer(contains("percent"), names_to = "housing_age") %>% 
  ggplot(aes(x = reorder(countyname, desc(countyname)), y = value, fill = housing_age)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = wes_palette(n = 2, name = "GrandBudapest1"),
                    name = "Housing Age",
                    labels = c("Built After 1960", "Built Before 1960")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title="Majority of the housing units are relatively new across most of the counties", subtitle="San Francisco and LA counties, however, have higher % of older housing units", x="Counties in California", y = "% Housing Units") +
  coord_flip() +
  theme_minimal()
ggsave(here("output", "housing_new_old.png"), width = 10, height = 10)
```

```{r eval=FALSE}
# Visualize housing structure percentages
# TODO: Fix mutate + across
df %>% 
  drop_na() %>% 
  select(countyname, contains("estimate_units_in_structure"), total_housing_units) %>% 
  # mutate(across(contains("estimate_units"), ~.x / total_housing_units, .names = "percent_{.col}"))
  mutate(across(contains("estimate_units"), ~ .data.col / .data$total_housing_units))
  
  xpivot_longer(contains("estimate_units_in_structure"), names_to = "units_in_structure") %>% 
  ggplot(aes(x = reorder(countyname, desc(countyname)), y = value, fill = units_in_structure)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_brewer(palette = "Dark2",
                    name = "Units in Structure",
                    labels = c("1 Unit Attached", "1 Unit Detached", "10 to 19 Units", "2 Units", "20 or More Units", "3 or 4 Units", "5 to 9 Units", "Boat, RV, Van, etc.", "Mobile Home")) +
  labs(title="Large variance in race and ethnicity makeup by county", subtitle="Percentage of each race and ethnicity group by county", x="Counties in California", y = "% Housing Units") +
  coord_flip() +
  theme_minimal()
ggsave(here("output", "population_race.png"), width = 10, height = 10)
```

The number of occupants per room within a house, however, is typically 1 or less across all counties. A few counties, e.g., Monterey, have a higher percentage of housing units where the occupants per room is greater than 1, signaling a more populated housing environment.

Note: Occupants per room columns are formatted as characters. 

```{r}
# Occupants per room
# TODO: Can I sort by "1 or Less" bar value?
df_occupancy <- df %>% 
  select(countyname, contains("estimate_occupants_per_room_occupied_housing_units")) %>% 
  mutate(across(contains("estimate_occupants_per_room_occupied_housing_units"), as.numeric)) %>% 
  drop_na() %>% 
  mutate(percent_1_00_or_less = estimate_occupants_per_room_occupied_housing_units_1_00_or_less / estimate_occupants_per_room_occupied_housing_units,
         percent_1_01_to_1_50 = estimate_occupants_per_room_occupied_housing_units_1_01_to_1_50 / estimate_occupants_per_room_occupied_housing_units,
         percent_1_51_or_more = estimate_occupants_per_room_occupied_housing_units_1_51_or_more / estimate_occupants_per_room_occupied_housing_units
         ) 

df_occupancy %>% 
  pivot_longer(contains("percent"), names_to = "occupancy_rate") %>% 
  ggplot(aes(x = reorder(countyname, desc(countyname)), y = value, fill = occupancy_rate)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = wes_palette(n = 5, name = "Darjeeling1"),
                    name = "Occupancy Rate",
                    labels = c("1 or Less", "1.01 to 1.50", "1.51 or More")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title="Majority have a 1 or less occupants per room ratio in all counties", subtitle="Some counties, e.g., Monterey, have a higher % of more than 1 occupants per room", x="Counties in California", y = "% Housing Units") +
  coord_flip() +
  theme_minimal()
ggsave(here("output", "housing_occupancy_rate.png"), width = 10, height = 10)
```

### COVID-19

Surprisingly, some of the top counties by COVID confirmed cases and deaths per 100K are not the most-populated counties. Furthermore, ranking of confimed cases and deaths did not always align, e.g, LA county had the 9th highest number of confirmed cases per 100K but 3rd highest number of deaths from COVID, signaling either disparity in access to testing or a challenge with treating confirmed patients.

Imperial county has an unusually high number of confirmed cases and deaths per 100K compared to other counties (LA Times article about Imperial County listing out possible factors: https://www.latimes.com/california/story/2020-07-28/imperial-county-coronavirus-rural-california).

```{r}
df_covid <- df %>% 
  select(countyname, deaths_per_100k = death_rate, cases_per_100k = confirmed_case_rate) %>% 
  drop_na()

df_covid %>% 
  ggplot(aes(x = reorder(countyname, cases_per_100k), y = cases_per_100k)) +
  geom_bar(stat = "identity", fill = wes_palette(n = 1, "GrandBudapest1")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title="COVID confirmed cases per 100K varies a lot between counties", subtitle="Top death counties and top confirmed cases counties are not exactly identical", x="Counties in California", y = " COVID Confirmed Cases per 100K People") +
  coord_flip() +
  theme_minimal()
ggsave(here("output", "covid_cases.png"), width = 10, height = 10)

df_covid %>% 
  ggplot(aes(x = reorder(countyname, deaths_per_100k), y = deaths_per_100k)) +
  geom_bar(stat = "identity", fill = wes_palette(n = 1, "GrandBudapest1")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title="COVID deaths per 100K varies a lot between counties", subtitle="Some had way more deaths per 100K than the others, even than the LA county", x="Counties in California", y = " COVID Deaths per 100K People") +
  coord_flip() +
  theme_minimal()
ggsave(here("output", "covid_deaths.png"), width = 10, height = 10)

df_covid %>% 
  ggplot(aes(x = cases_per_100k, y = deaths_per_100k)) +
  geom_point() +
  labs(title="COVID deaths counts per 100K were correlated with the confirmed cases counts", subtitle="Some counties, however, observed a lower or higher death rate per case compared to other counties", x="Confimred COVID Cases per 100K People", y = " COVID Deaths per 100K People") +
  theme_minimal()
ggsave(here("output", "covid_cases_deaths.png"), width = 10, height = 10)

```

## Relationship between variables

Hypothesis for association:

Population vs. housing units
Population vs. median rooms
Population vs. estimate_occupants_per_room_occupied_housing_units
Population vs. COVID

```{r}
# Note: Can I get the housing data by race and ethnicity so I can color-code the plots by the race / ethniciy group?
df %>% 
  select(population = county_pop2019, 
         number_rooms = estimate_rooms_total_housing_units_median_rooms, 
         covid_deaths = death_rate, 
         covid_cases = confirmed_case_rate) %>% 
  drop_na() %>% 
  ggpairs()
```

```{r}
df_selected <- df %>% 
  select(county = countyname,
         population = county_pop2019,
         parks_usage = parks_usage_percent_during_covid,
         workspace_usage = workspace_usage_percent_during_covid,
         residential_usage = residential_usage_percent_during_covid,
         multi_unit_housing = housing_units_in_multi_unit_structures_percent,
         pp_household = persons_per_household,
         income = median_household_income_in_dollar,
         race = dominant_race_y,
         pct_white = percent_white,
         ethnicity = dominant_ethnicity_y,
         covid_deaths = death_rate
         ) %>% 
  drop_na() %>% 
  mutate(race = if_else(str_detect(race, "non-white"), "Majority non-white", "Majority white")
  )

colnames(df_selected)
```

Variables Associated with COVID Death Rate

* Persons per household have a positive correlation with the COVID death rate (i.e., counties with households of more people are associated with a higher COVID death rate). This could be because families can spread the disease among themselves.
* Percentage of white demograhpics within the county population has a negative correlation with the COVID death rate. This may indicate majority white counties are more likely to be better equipped to deal with COVID infection.
* The increase in park usage since COVID is negatively correlated with the COVID death rate. This could indicate 1) counties with more parks have more resources to deal with COVID infection or more people who are active; 2) Outdoor time spent has a positive association with a lower COVID death rate either due to its positive impact on people's health (e.g., getting fresh air or preventing vitamin D defficiency); or 3) counties with a lower COVID death rate are more likely to spend time outside of their homes.
* As we suspected, population is not correlated with the COVID death rate, suggesting there are disparities among counties that affected the COVID death rate.
* Unlike our hypothesis, lower income, higher precentage of multi-unit housing (i.e., more chance of interacting with others in the building), lower residential usage/higher workspace usage did not have a positive associate with the COVID death rate.


```{r corr_plot}
corr <- round(cor(df_selected %>% select(is.numeric) %>% rename_with(~ str_replace_all(.x, "_", " ") %>% str_to_title())), 1)
corr

# Compute a matrix of correlation p-values
p.mat <- cor_pmat(df_selected %>% select(is.numeric))
p.mat
  
# Visualize with correlation coefficients
# Add correlation significance level (X if not significant)
ggcorrplot(corr,
           type = "lower",
           outline.color = "white",
           colors = c("#6D9EC1", "white", "#E46726"),
           lab = TRUE,
           p.mat = p.mat,
           insig = "blank",
           sig.level = 0.05) +
  labs(title = "Pearson correlation matrix among selected variables", subtitle="Shown only if the correlation p-value is less than 0.05")
ggsave(here("output", "corr_matrix.png"), width = 10, height = 10)
```

```{r plot_race_ethnicity}
df_selected %>% 
  select(race, covid_deaths) %>% 
  mutate(race = if_else(str_detect(race, "non-white"), "Majority non-white (n = 18)", "Majority white (n = 40)")) %>% 
  ggplot(aes(x = race, y = covid_deaths / 100)) + 
  geom_jitter(width = 0.1) +
  stat_summary(fun.y=median, geom="point", shape=18,
                 size=3, color="red") +
  scale_y_continuous(labels = scales::percent) +
  # labs(title="Counties with majority non-whte population have a higher COVID death rate", subtitle="(Median values highlighted in red)", x="Dominant Population Within County", y = "COVID Death Rate per 100K People") +
  # For the presentation deck
  labs(title="COVID Death Rates by Race", subtitle="(Median values highlighted in red)", x="Largest Population Within County", y = "COVID Death Rate per 100K People") +
  theme(text = element_text(size=15)) +
  theme_minimal()
  
ggsave(here("output", "covid_race.png"), width = 10, height = 10)

df_selected %>% 
  select(ethnicity, covid_deaths) %>% 
  ggplot(aes(x = ethnicity, y = covid_deaths / 100)) + 
  geom_jitter(width = 0.1) +
  stat_summary(fun.y=median, geom="point", shape=18,
                 size=3, color="red") +
  scale_y_continuous(labels = scales::percent) +
  labs(title="Counties with majority Hispanic or Latino population have a higher COVID death rate", subtitle="(Median values highlighted in red)", x="Dominant Population Within County", y = "COVID Death Rate per 100K People") +
  theme_minimal()
ggsave(here("output", "covid_ethnicity.png"), width = 10, height = 10)
```

```{r}
df_selected %>% 
  arrange(desc(covid_deaths)) %>% 
  head(10) %>% 
  select(county, race, covid_deaths_per_100k = covid_deaths)
```

In the correlation matrix, we saw the persons per household had a negative association with the COVID death rate. After fitting a larger logistic model with all independent variables that were available, we found that this trend was introduced via confounding effects of other variables such as household income.

```{r fit_lm}
# Fit a linear regression model: Association "after" accounting for other variables
lm.fit = lm(covid_deaths ~., data = df_selected %>% select(-county))
summary(lm.fit)

lm.fit = lm(covid_deaths ~ race, data = df_selected)
summary(lm.fit)

lm.fit = lm(covid_deaths ~ race + pct_white, data = df_selected)
summary(lm.fit)

lm.fit = lm(covid_deaths ~ pct_white + income, data = df_selected)
summary(lm.fit)

lm.fit = lm(covid_deaths ~ pct_white + income + parks_usage, data = df_selected)
summary(lm.fit)

lm.fit = lm(covid_deaths ~ income + parks_usage, data = df_selected)
summary(lm.fit)

lm.fit = lm(covid_deaths ~ race + income + parks_usage, data = df_selected)
summary(lm.fit)

lm.fit = lm(covid_deaths ~ ethnicity + income + parks_usage, data = df_selected)
summary(lm.fit)

lm.fit = lm(covid_deaths ~ race + income + parks_usage + pp_household, data = df_selected)
summary(lm.fit)

```


